---
title: "Identifying a cutoff value from bimodal data"
csl: the-american-naturalist.csl
output:
  html_document:
    theme: cerulean
    toc: yes
bibliography: references.bib
---

```{r include=FALSE}
knitr::knit_hooks$set(margin = function(before,options,envir) {
  if(before) par(mgp=c(1.5,0.5,0),bty="n",plt=c(.105,1,.13,1))
  else NULL
})
knitr::opts_chunk$set(margin=T,prompt=T,comment="",collapse=T,cache=T,
                      dev.args=list(pointsize=11),fig.height=3.5,
                      fig.width=4.24725,fig.retina=2)
#hist <- function(x,...) graphics::hist(x,ylim=c(0,.4),main=NULL,col="grey",...)
hash <- c(normal=dnorm,"log-normal"=dlnorm,gamma=dgamma,weibull=dweibull)
```


This tutorial shows step by step how to use finite mixture models to model
bimodal data and to derive a cut-off value that separates the two peaks for
a given type-I error. It shows in particular how to estimate the parameters of
the finite mixture model with the Expectation-Maximization (E-M) algorithm
together with the confidence intervals of the estimations. This method is used
in [[@Trang2015]](www.rstudio.com). For questions and/or help, contact me at
marc.choisy@ird.fr.

### Installing and loading the \code{cutoff} package

Once you've download the package \code{cutoff} from [here](www.rstudio.com) you
can install it by the following command:
```{r eval=FALSE}
install.packages("path_of_the_package/cutoff-0.1.0.tar.gz")
```
This needs to be done only once on a computer, unless there is a new version of
R or the package.

Once the package is installed on your computer you need to load it to use it
```{r}
library(cutoff)
```
and this need to be done for each working session. This package contains two
main functions:
* \code{em} that fit a finite mixture model to bimodal data with the
Expectation-Maximization algorithm
* \code{cutoff} that calculate a cutoff value from a fitted finite mixture
model, given a type-1 error
There are additional functions, mainly to visualize the results and that we
introduce below.

### Bimodal data

The data should be contained in a numerical vector. Here is an example of
bimodal data containing the IgG concentration against measles virus:
```{r}
length(measles)
range(measles)
# A histogram of the data:
hist(measles,100,F,xlab="concentration",ylab="density",ylim=c(0,.55),
     main=NULL,col="grey")
# A kernel density estimation of the distribution:
lines(density(measles),lwd=1.5,col="blue")
```

This figure shows the histogram of the data together with a non-parametric
estimation of the distribution. The two suggest a bimodal distribution of
the data.

### Finite mixture models

We can model such a bimodal distribution by a finite mixture model that uses
continuous distributions from the exponential family [@Schlattmann2009]. In
absence of clear expectation on the shape of these distributions, we can
consider the normal, the gamma and the Weibull distributions. The normal
distribution is defined on all the real numbers, whereas the gamma and Weibull
distributions are defined on positive reals. These three distributions are
characterized by 2 parameters: a location parameter $\mu$ and a scale parameter
$\sigma$. The first parameter accounts for the location of most of the data and
corresponds to the mean of the normal distribution, and the shape parameter for
the gamma and Weibull distributions. The second parameter accounts for the
spread of the data around the location parameter and corresponds to the standard
deviation in the case of the normal distribution, the rate parameter (1/scale)
for the gamma distribution and the scale parameter for the Weibull distribution.
We refer to $\mu_1$ and $\sigma_1$ for the location and scale parameters of the
peak of the lower concentrations and $\mu_2$ and $\sigma_2$ for the location and
scale parameters of the peak of the higher concentrations. The density of the
bimodal distribution of concentrations thus reads
$$
f(x|\lambda,\mu_1,\sigma_1,\mu_2,\sigma_2) =
  \lambda\times\mathcal{D}_1(x|\mu_1,\sigma_1) +
  (1-\lambda)\times\mathcal{D}_2(x|\mu_2,\sigma_2)
$$

The parameters of the finite mixture model are estimated by the E-M algorithm
[@Do2008] as coded by the \code{em} function. These parameters include two
parameters for each of the two probability distributions and a mixture parameter.
```{r}
# Estimating the parameters of the finite mixture model:
(measles_out <- em(measles,"normal","normal"))
# The confidence interval of the parameter estimates:
confint(measles_out,nb=100,level=.95)
# The plot:
hist(measles,100,F,xlab="concentration",ylab="density",ylim=c(0,.55),
     main=NULL,col="grey")
lines(measles_out,lwd=1.5,col="red")
```

Confidence interval for the mixture parameter $\lambda$ is found using the
method of @Oakes1999.

### Identifying a cutoff value

We can use the fitted finite mixture model to identify a cutoff value that
discriminate the two modes of the dataset. For that, we compute the probability
for a datum to belong to distribution $\mathcal{D}_1$ as
$$
p_1 = \frac{\lambda\times\mathcal{D}_1(x|\mu_1,\sigma_1)}
  {\lambda\times\mathcal{D}_1(x|\mu_1,\sigma_1)+
  (1-\lambda)\times\mathcal{D}_2(x|\mu_2,\sigma_2)}
$$
and the probability for a datum to belong to distribution $\mathcal{D}_1$ as
$$
p_2 = \frac{(1-\lambda)\times\mathcal{D}_2(x|\mu_2,\sigma_2)}
  {\lambda\times\mathcal{D}_1(x|\mu_1,\sigma_1)+
  (1-\lambda)\times\mathcal{D}_2(x|\mu_2,\sigma_2)}
$$
We equate this probability to the type-I error we aim at to find the cutoff
value. The confidence interval of the cutoff value is computed by Monte Carlo
simulations (see the help of the `cutoff` function for more details). In
practice, this gives:
```{r}
hist(measles,100,F,xlab="concentration",ylab="density",ylim=c(0,.55),
     main=NULL,col="grey")
lines(measles_out,lwd=1.5,col="red")
# Estimating a cutoff value from this fitted finite mixture model:
(cut_off <- cutoff(measles_out))
# Plotting it:
polygon(c(cut_off[-1],rev(cut_off[-1])),c(0,0,.55,.55),
        col=rgb(0,0,1,.2),border=NA)
abline(v=cut_off[-1],lty=2,col="blue")
abline(v=cut_off[1],col="blue")
```


### References